Coherence between subjective feeling and facial expression: A comparison of human and automatic recognition

### 2019-04-15 ACII_2019_paper/ACII_2019_paper.tex
new version

% Pandoc header
\usepackage{booktabs}
\usepackage{float}
\usepackage{tabu}
\usepackage{wrapfig}
\floatplacement{figure}{H}
\floatplacement{table}{H}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

With the development of commercial automatic facial expression
recognition tools (see Dupré et al. 2018 for a non-exhaustive list of
available tools), industries and governments are gradually implementing
this technology in order to track humans' emotions in various scenarios
(e.g., marketing, healthcare, automotive to name a few). This
technoloogy rests on the premise that facial expressions provide a
direct access to individuals' subjective feeling. Even if this premise
is central to the modern mainstream approach of human emotion, recent
research in affective science are challenging it. Once the two competing
approaches briefly described as well as the predictions they
respectively entail, an experiment testing these hypotheses will be
presented and its results analyzed in order to provide empirical
evidence to contribute to answer the question.

\hypertarget{the-link-between-emotion-and-facial-expression-through-the-behaviorist-approach}{%
\subsection{The link between emotion and facial expression through the
behaviorist
approach}\label{the-link-between-emotion-and-facial-expression-through-the-behaviorist-approach}}

Based on the behaviorist approach initiated by Darwin in \emph{The
Expression of the Emotions in Man and Animals} (Darwin 1872), facial
expressions are conceived as a genuine displays of individuals' inner
emotional state. This hypothesis is used as a basis for the Basic
Emotion Theory (BET) which states that a set of six emotions are
universally displayed and are genetically hardwired not only in humans
(Ekman 1992) but also in different animal species (Waal 2019). According
to this view, ``when emotions are aroused by perception of a social
event, a set of central commands produce patterned emotion-specific
changes in multiple systems, including {[}\ldots{}{]} facial
expressions.'' (Ekman 2007, p49). To cope with critics, several
amendments have been made to the BET, increasing the number of basic
emotions from six to seven (Ekman and Heider 1988) as well as adding the
concept of ``display rules'' to explain cultural differences in the
management of facial expressions (Ekman et al. 1987).

Even if this theory obtained a popular support, it fails to explain how
individuals can feel emotions without expressing them and how
individuals can express emotions without feeling them, in instances in
which display rules cannot be called upon (Kraut and Johnston 1979;
Durán, Reisenzein, and Fernández-Dols 2017). These evidences have led to
an alternative conception, notably the social constructivist approach.

\hypertarget{the-link-between-emotion-and-facial-expression-through-the-social-constructivist-approach}{%
\subsection{The link between emotion and facial expression through the
social constructivist
approach}\label{the-link-between-emotion-and-facial-expression-through-the-social-constructivist-approach}}

Detractors of the Basic Emotion Theory consider emotion not as
genetically hardwired but as a learnt association between a given
situation and an appropriate response (Averill 1980; Barrett 2017a). For
the tenants of the constructivist approach, emotions are ``concepts''
based on past experiences and which are ``a collection of embodied,
whole brain representations that predict what is about to happen in the
sensory environment, what the best action is to deal with impending
events, and their consequences for allostasis'' (Barrett 2017b, p12).
Following this assumption, faces are best conceived as tools displaying
signals in social interactions (Crivelli and Fridlund 2018). These
signals can convey individuals' motivations and readiness (Frijda and
Tcherkassof 1997) or social messages (Fridlund 1994). Therefore, facial
expressions are thought as behaviors which meaning is inferred by the
observer. Findings support this observer dependence (Lindquist and
Gendron 2013; Niedenthal et al. 2017). They show that to make meaning of
another person's facial behavior, the perceiver relies in particular on
her/his knowledge about emotion categories.

\hypertarget{the-recognition-of-emotional-facial-expressions}{%
\subsection{The recognition of emotional facial
expressions}\label{the-recognition-of-emotional-facial-expressions}}

Regarding emotional facial expression (EFE) recognition, the behaviorist
approach assertions and the constructivist approach claims lead to two
divergent predictions. The foremost postulates that, when triggered,
each basic emotion is expressed by a prototypical face (non basic
emotions being blends of the basic ones). Basic emotions are easily
recognized by all human observers and emotional states are accessible by
facial measurement. As the recognition of facial expressions is based on
the identification of specific patterns of facial movements, it implies
that EFE recognition of both human observers and automatic classifiers
should be as accurate. The constructivist approach, in contrast, affirms
that facial expressions do not provide a direct access to individuals'
subjective feeling. Even if the face possibly moves during an emotional
episode, facial muscle movements are not linked in a one-to-one manner
to a specific discrete emotional experience (Doyle and Lindquist 2017,
p418). Instead, emotions are mentally constructed by the perceiver and
mental categories of emotions are needed to accurately categorize facial
movements. Therefore, since there is no emotional prototypical face, one
should expect human observers' superior ability to accurately recognize
EFE as compared to automatic EFE recognition tools.

@@

\hypertarget{method}{%
\section{Method}\label{method}}

@@

\hypertarget{emotion-elicitation}{%
\subsection{Emotion Elicitation}\label{emotion-elicitation}}

For the emotion elicitation experiment, 358 French participants (182
females, 176 males, \emph{M}age = 47.9, \emph{SD}age = 9.2) were
recruited to perform one out of 11 emotion elicitation tasks designed to
trigger a positive, a specific negative or a neutral emotional state.
Encoders' face were recorded using an hidden camera resulting 358 front
facing 768x576 videos varying from 1s to 1479s (Figure
\ref{fig:dynemo_img}). These recordings form the DynEmo database (see
Tcherkassof et al. 2013 for a full description of tasks and procedure).

@@

After the emotion elicitation task the encoders rated their subjective
feeling on Likert scales from 0 (``not at all'') to \nolinebreak 5
\nolinebreak (``strongly'') related to six ``basic'' emotion labels
(i.e., \emph{anger}, \emph{disgust}, \emph{fear}, \emph{happiness},
\emph{surprise} and \emph{sadness}) as well as six ``non-basic'' emotion
labels (i.e., \emph{pride}, \emph{curiosity}, \emph{boredom},
\emph{shame}, \emph{humiliation}, and \emph{disappointment}).

Finally, a debriefing session was performed to ensure that encoders were
not durably affected by the emotion elicitation task. The debriefing was
also used to check that encoders did not guess the real purpose of the
experiment (e.g.~being filmed while they were performing an emotional
elicitation task) to guarantee facial expressions' genuineness. All
encoders gave their agreement on their data and video to be processed
for research purpose only.

\hypertarget{human-facial-expression-recognition}{%
\subsection{Human Facial Expression
Recognition}\label{human-facial-expression-recognition}}

@@

The annotation of facial expressions was performed on-site using
\emph{Oudjat}, a software for designing video annotation experiments
(Dupré et al. 2015). For each video, the annotation procedure followed
two steps. First, the participants had to identify the emotional
sequences by pressing the space bar of their keyboard to indicate the
beginning and the end of the emotional sequences while watching the
video. Second, the participants watched each emotional sequence
previously identified and labeled the sequence using one of the 12
\nolinebreak emotions proposed including six ``basic'' emotion labels
(i.e., \emph{anger}, \emph{disgust}, \emph{fear}, \emph{happiness},
\emph{surprise} and \emph{sadness}) and six ``non-basic'' emotion labels
(i.e., \emph{pride}, \emph{curiosity}, \emph{boredom}, \emph{shame},
\emph{humiliation}, and \emph{disappointment}). They also had the
possibility to indicate that the sequence was expressing none of the
proposed emotion labels.

@@

\hypertarget{automatic-facial-expression-recognition}{%
\subsection{Automatic Facial Expression
Recognition}\label{automatic-facial-expression-recognition}}

The 232 annotated video were processed with Affdex (SDK v3.4.1). Affdex
is an automatic facial expression recognition classifier developed and
distributed by Affectiva is a spin-off company resulting from the
research activities of MIT media lab created in 2009 (McDuff et al.
2016). Affdex's algorithm uses Histogram of Oriented Gradient (HOG)
features and Support Vector Machine (SVM) classifiers in order to
recognize facial expressions. For each video frame, Affdex identifies
the probability \(p_{video_{i}.label_{j}.t_{k}}\) from 0 to 100
(rescaled to 0 to 1 for the analysis) of the face as expressing one of
the six ``basic'' emotion labels (i.e., \emph{anger}, \emph{disgust},
\emph{fear}, \emph{happiness}, \emph{surprise} and \emph{sadness}) as
well as additional psychological states such as \emph{valence},
\emph{engagement} or \emph{contempt}, and facial features such as
\emph{cheek raise}, \emph{eye widen} or \emph{jaw drop}.

For both human and automatic recognition, to determine which of the six
``basic'' emotions can be used to identify each video, the recognition
probability for each label by frame was converted into odd ratio by
label (Dente et al. 2017). The highest sum of each odd ratio time-series
defines the label recognized by the automatic classifier (EQ\ref{eq:2}
for human recognition and EQ\ref{eq:3} for automatic recognition).
\begin{equation}
\label{eq:2}
video_{i}.label = \max\left(\frac{\sum_{k=1}^{n}x_{video_{i}.label_{j}.t_{k}}}{\sum_{k=1}^{n}x_{video_{i}.t_{k}}}\right)
\end{equation}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\hypertarget{self-reported-emotions-recognized-by-human-observers}{%
\subsection{Self-reported emotions recognized by human
observers}\label{self-reported-emotions-recognized-by-human-observers}}

Encoders' subjective feeling is compared with human observers
recognition in a confusion matrix (Figure
\ref{fig:confusionMatrix_sr_hr}). Each emotion label used to describe
encoders' self-reported subjective feeling (i.e., the label rated with
the highest value) is compared with the emotion labels which were rated
with the highest score by human observers.

@@

Results of the confusion matrix show a moderate agreement between the
emotion felt by the encoder during the elicitation and the emotion
recognized by the human observers (Accuracy \nolinebreak =
\nolinebreak 0.43, 95\%CI{[}0.35,0.52{]}; Kappa = 0.19) except for
\emph{disgust} (100\% of the videos self-reported). Sensitivity,
specificity, precision and F1 score for each emotion reveals that
\emph{happiness} has the highest coherence ratio whereas \emph{sadness}
has the lowest coherence ratio between true positives and false
positives (Table \ref{table:confusionTable_sr_hr}). Interestingly human
observers seem to recognize \emph{surprise} expressed in videos where
\emph{anger}, \emph{fear} \emph{happiness} and \emph{sadness} was the
highest self-reported emotion (respectively 28.6\%, 33.3\%, 34.7\% and
33.3\% of the videos self-reported), and in a lower instance
\emph{happiness} was recognized in videos where \emph{fear} and
\emph{surprise} was the highest self-reported emotion (respectively
50.0\% and 25.6\% of the videos self-reported).

@@

@@

\hypertarget{self-reported-emotions-recognized-by-the-automatic-classifier}{%
\subsection{Self-reported emotions recognized by the automatic
classifier}\label{self-reported-emotions-recognized-by-the-automatic-classifier}}

Similarly to the previous analysis, a confusion matrix was used to
compare encoders' subjective feeling with the emotion label recognized
by the automatic classifier (Figure \ref{fig:confusionMatrix_sr_ar}).

@@

Results obtained for the comparison between emotions self-reported and
recognized by the automatic classifier are somewhat similar to the ones
with human observers (Table \ref{table:confusionTable_sr_ar}). Overall,
a low agreement between emotion self-reported and emotion recognized by
the automatic classifier (Accuracy \nolinebreak = \nolinebreak 0.3,
95\%CI{[}0.22,0.38{]}; Kappa = 0.07) except for \emph{happiness} (44.4\%
of the video self-reported) is watched. Surprisingly the automatic
classifier incorrectly recognized as \emph{disgust} an important
proportion of videos in which \emph{anger}, \emph{happiness} and
\emph{surprise} was the highest self-reported emotion (respectively
42.9\%, 25.0\% and 32.6\% of the videos self-reported). In parallel, the
automatic classifier recognized as \emph{happiness} videos in which
\emph{fear} and \emph{surprise} was the highest self-reported emotion
(respectively 66.7\% and 32.6\% of the videos self-reported).

\begin{table}[!h]

@@

\hypertarget{comparison-between-human-and-automatic-recognition}{%
\subsection{Comparison between human and automatic
recognition}\label{comparison-between-human-and-automatic-recognition}}

As previously mentioned, human observers appear to be more accurate than
the automatic classifier to recognize individuals' subjective feeling
(human observer Accuracy = 0.43; automatic classifier Accuracy = 0.3).
However, both make mistakes. A third confusion matrix is used to compare
similarities (diagonal) and differences between human observers and
automatic classifier in classifing the six emotion labels (Figure
\nolinebreak \ref{fig:confusionMatrix_hr_ar}).

@@

\hypertarget{discussion-and-conclusion}{%
\section{Discussion and Conclusion}\label{discussion-and-conclusion}}

@@

Such results support the hypothesis advanced by some authors of a low
emotion--expression coherence (e.g., Kappas 2003). In many instances,
facial displays are not associated with a concordant emotional state,
even any emotional state at all (Bonanno and Keltner 2004;
Fernández-Dols and Crivelli 2013). More and more evidences are showing
that facial expressions are in reality not expressing emotions (McKeown
2013). As well as other nonverbal behaviors, facial movements are not
only assumed to be determined by emotion but also by various other
causes, such as psychological states (e.g., \nolinebreak motivations or
pain), to say nothing of social context and sociocultural norms (such as
``display rules'', Ekman et al. 1987). This multiple determination
excludes any possibility of drawing an inference from facial activity on
the underlying psychological state (emotional or other). Beyond the
present observations showing a weak coherence between subjective
feelings and spontaneous facial expressions, this study provides some
light on the controversy between the behaviorist and the constructivist
approaches. The first of those two lines of thinking (Ekman 1992)
assumes that expressions of emotion are brief and coherent patterns of
facial muscle movements that co-vary with discrete subjective
experiences. Emotions and their related prototypical facial behavior are
universal because they are considered as innate mechanisms allowing
individuals to respond adaptively to evolutionary significant events
(threats, opportunities\ldots{}) encountered in the environment. Instead
of viewing emotions as natural kinds (Barrett 2017b), the constructivist
approach supposes that emotions are social constructions. The specific
emotion categories identified on the face of another person are the
result of categorization. In other words, the emotions that are
recognized by the observer are constructed in her/his mind, primarily
based on her/his conceptual knowledge of emotion. Therefore, facial
movements do not express specific emotions. It is the observer that
infers the emotional meaning of the facial expression. As a consequence,
one can predict from the first line of thinking that individuals'
emotional subjective feeling should be correlated to the recognition of
facial expressions from both human observers and automatic classifiers
whereas if emotions are social constructs, as stated by the second line
of thinking, human observer's should be better at perceiving emotions in
face than automatic classifiers. Present results plead in favor of the
latter stance. They show that human observers are more accurate than the
automatic recognition tool to identify individuals' subjective feeling
on the basis of their face. Moreover, mistakes made by human observers
look less arbitrary to the ones made by the classifier. For instance,
even if mix-up between disgust and anger is sometimes reported in
recognition studies, confusions such as the ones produced by the
classifier have never been noted for human observers. A possible
explanation is that human observers are assessing dynamic expressive
sequences. They have access to the full video which provides a kinetic
context whereas the automatic classifier only assesses the videos frame
by frame. Further research is needed to shed light on this issue.

Some limitations should be stated, notably regarding the use of
self-reports to evaluate encoders' subjective feelings. Accessing the
inner subjective feeling can be biased if not impossible. Moreover the
procedure used for human observation can also be open to dispute.
Instead of asking the human annotators to provide a unique label, a more
subtle approach was chosen to mimic results provided by the automatic
classifier . Whereas this paradigm is longer and more complicated, it
can lead to more robust results in reducing the forced-choice bais
(Russell 1993). However this procedure can also reduce the human
observers accuracy. In this regard, the results of the human observation
could have been more ambiguous because it is not the natural way that
people are inferring meaning from facial expressions. An alternative
explanation relies in reducing the recognition bais involved in the
classic recognition paradigm. Classic forced-choice paradigms obtain
artificial high results, thus by using a more evolved approach
observers' accuracy may have been lowered.

Considering the above, the results provide additional evidence that
individuals' subjective feeling can not be inferred from facial
expressions and in our case invalidate the hypothesis of hardwired
emotions unambiguously displayed on the face. Even if emotions were
hardwired, in everyday life one does not observe prototypical facial
expressions and therefore research should be focused on analysing
non-prototypical facial expressions. Advancements in identifying
``non-basic'' emotion labels (McDuff 2016) as well as non-prototypical
facial expression have been made in the developpement of automatic
facial expression recognition tools. However this result suggests that
automatic facial expression recognition tools should merely evaluate
facial morphology features such as action units (e.g., OpenFace
presented in Baltrušaitis, Robinson, and Morency 2016) rather than
inferring supposedly emotional or affective states.

\hypertarget{acknowledgment}{%
\section{Acknowledgment}\label{acknowledgment}}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-averill1980constructivist}{}%
Averill, James R. 1980. ``A Constructivist View of Emotion.'' In
\emph{Theories of Emotion}, edited by Robert Plutchik and Henry
Kellerman, 305--39. New York, NY: Academic Press.

>>>>

% Pandoc header
\usepackage{booktabs}
\usepackage{float}
\usepackage{tabu}
\usepackage{wrapfig}
\usepackage[none]{hyphenat}
\floatplacement{figure}{H}

@@

\begin{IEEEkeywords}
Facial expression, self-report, human observer, automatic recognition.
\end{IEEEkeywords}

\hypertarget{i.-introduction}{%
\section{I. Introduction}\label{i.-introduction}}

With the development of commercial automatic facial expression
recognition tools {[}1{]}, industries and governments are gradually
implementing this technology in order to track humans' emotions in
various scenarios (\emph{e.g.}, marketing, healthcare, automotive to
name a few). This technoloogy rests on the premise that facial
expressions provide a direct access to individuals' subjective feeling.
Even if this premise is central to the modern mainstream approach of
human emotion, recent research in affective science are challenging it.
Once the two competing approaches briefly described as well as the
predictions they respectively entail, an experiment testing these
hypotheses will be presented and its results analyzed in order to
provide empirical evidence to contribute to answer the question.

\hypertarget{a.-the-behaviorist-approach}{%
\subsection{A. The Behaviorist
Approach}\label{a.-the-behaviorist-approach}}

Based on the behaviorist approach initiated by Darwin in \emph{The
Expression of the Emotions in Man and Animals} {[}2{]}, facial
expressions are conceived as a genuine displays of individuals' inner
emotional state. This hypothesis is used as a basis for the Basic
Emotion Theory (BET) which states that a set of six emotions are
universally displayed and are genetically hardwired not only in humans
{[}3{]} but also in different animal species {[}4{]}. According to this
view, \emph{``when emotions are aroused by perception of a social event,
a set of central commands produce patterned emotion-specific changes in
multiple systems, including {[}\ldots{}{]} facial expressions.''} {[}5,
p. 49{]}. To cope with critics, several amendments have been made to the
BET, increasing the number of basic emotions from six to seven {[}6{]}
as well as adding the concept of ``display rules'' to explain cultural
differences in the management of facial expressions {[}7{]}.

Even if this theory obtained a popular support, it fails to explain how
individuals can feel emotions without expressing them and how
individuals can express emotions without feeling them, in instances in
which display rules cannot be called upon {[}8{]}, {[}9{]}. These
evidences have led to an alternative conception, notably the social
constructivist approach.

\hypertarget{b.-the-social-constructivist-approach}{%
\subsection{B. The Social Constructivist
Approach}\label{b.-the-social-constructivist-approach}}

Detractors of the Basic Emotion Theory consider emotion not as
genetically hardwired but as a learnt association between a given
situation and an appropriate response {[}10{]}, {[}11{]}. For the
tenants of the constructivist approach, emotions are ``concepts'' based
on past experiences and which are \emph{``a collection of embodied,
whole brain representations that predict what is about to happen in the
sensory environment, what the best action is to deal with impending
events, and their consequences for allostasis''} {[}12, p. 12{]}.
Following this assumption, faces are best conceived as tools displaying
signals in social interactions {[}13{]}. These signals can convey
individuals' motivations and readiness {[}14{]} or social messages
{[}15{]}. Therefore, facial expressions are thought as behaviors which
meaning is inferred by the observer. Findings support this observer
dependence {[}16{]}, {[}17{]}. They show that to make meaning of another
person's facial behavior, the perceiver relies in particular on her/his
knowledge about emotion categories.

\hypertarget{c.-emotional-facial-expression-recognition}{%
\subsection{C. Emotional Facial Expression
Recognition}\label{c.-emotional-facial-expression-recognition}}

Regarding emotional facial expression (EFE) recognition, the behaviorist
approach assertions and the constructivist approach claims lead to two
divergent predictions. The foremost postulates that, when triggered,
each basic emotion is expressed by a prototypical face (non basic
emotions being blends of the basic ones). Basic emotions are easily
recognized by all human observers and emotional states are accessible by
facial measurement. As the recognition of facial expressions is based on
the identification of specific patterns of facial movements, it implies
that EFE recognition of both human observers and automatic classifiers
should be as accurate. The constructivist approach, in contrast, affirms
that facial expressions do not provide a direct access to individuals'
subjective feeling. Even if the face possibly moves during an emotional
episode, facial muscle movements are not linked in a one-to-one manner
to a specific discrete emotional experience {[}18, p. p418{]}. Instead,
emotions are mentally constructed by the perceiver and mental categories
of emotions are needed to accurately categorize facial movements.
Therefore, since there is no emotional prototypical face, one should
expect human observers' superior ability to accurately recognize EFE as
compared to automatic EFE recognition tools (Figure
\ref{fig:models_img}).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/models_img-1.pdf}
\caption{\label{fig:models_img}Comparison of behaviorist and social
constructivist approaches according to the expected differences in
emotion recognition.}
\end{figure}

@@

\hypertarget{ii.-method}{%
\section{II. Method}\label{ii.-method}}

@@

\hypertarget{a.-emotion-elicitation}{%
\subsection{A. Emotion Elicitation}\label{a.-emotion-elicitation}}

For the emotion elicitation experiment, 358 French participants (182
females, 176 males, \emph{M}\textsubscript{age} = 47.9,
\emph{SD}\textsubscript{age} = 9.2) were recruited to perform one out of
11 emotion elicitation tasks designed to trigger a positive, a specific
negative or a neutral emotional state. Encoders' face were recorded
using an hidden camera resulting 358 front facing 768x576 videos varying
from 1s to 1479s (Figure \ref{fig:dynemo_img}). These recordings form
the DynEmo database {[}19{]}.

@@

After the emotion elicitation task the encoders rated their subjective
feeling on Likert scales from 0 (``not at all'') to \nolinebreak 5
\nolinebreak (``strongly'') related to six ``basic'' emotion labels
(\emph{i.e.}, \emph{anger}, \emph{disgust}, \emph{fear},
\emph{happiness}, \emph{surprise} and \emph{sadness}) as well as six
``non-basic'' emotion labels (\emph{i.e.}, \emph{pride},
\emph{curiosity}, \emph{boredom}, \emph{shame}, \emph{humiliation}, and
\emph{disappointment}).

Finally, a debriefing session was performed to ensure that encoders were
not durably affected by the emotion elicitation task. The debriefing was
also used to check that encoders did not guess the real purpose of the
experiment (\emph{e.g.}, being filmed while they were performing an
emotional elicitation task) to guarantee facial expressions'
genuineness. All encoders gave their agreement on their data and video
to be processed for research purpose only.

\hypertarget{b.-human-facial-expression-recognition}{%
\subsection{B. Human Facial Expression
Recognition}\label{b.-human-facial-expression-recognition}}

@@

The annotation of facial expressions was performed on-site using
\emph{Oudjat}, a software for designing video annotation experiments
{[}20{]}. For each video, the annotation procedure followed two steps.
First, the participants had to identify the emotional sequences by
pressing the space bar of their keyboard to indicate the beginning and
the end of the emotional sequences while watching the video. Second, the
participants watched each emotional sequence previously identified and
labeled the sequence using one of the 12 \nolinebreak emotions proposed
including six ``basic'' emotion labels (\emph{i.e.}, \emph{anger},
\emph{disgust}, \emph{fear}, \emph{happiness}, \emph{surprise} and
\emph{sadness}) and six ``non-basic'' emotion labels (\emph{i.e.},
\emph{pride}, \emph{curiosity}, \emph{boredom}, \emph{shame},
\emph{humiliation}, and \emph{disappointment}). They also had the
possibility to indicate that the sequence was expressing none of the
proposed emotion labels.

@@

\hypertarget{c.-automatic-facial-expression-recognition}{%
\subsection{C. Automatic Facial Expression
Recognition}\label{c.-automatic-facial-expression-recognition}}

The 232 annotated video were processed with Affdex (SDK v3.4.1). Affdex
is an automatic facial expression recognition classifier developed and
distributed by Affectiva is a spin-off company resulting from the
research activities of MIT media lab created in 2009 {[}21{]}. Affdex's
algorithm uses Histogram of Oriented Gradient (HOG) features and Support
Vector Machine (SVM) classifiers in order to recognize facial
expressions. For each video frame, Affdex identifies the probability
\(p_{video_{i}.label_{j}.t_{k}}\) from 0 to 100 (rescaled to 0 to 1 for
the analysis) of the face as expressing one of the six ``basic'' emotion
labels (\emph{i.e.}, \emph{anger}, \emph{disgust}, \emph{fear},
\emph{happiness}, \emph{surprise} and \emph{sadness}) as well as
additional psychological states such as \emph{valence},
\emph{engagement} or \emph{contempt}, and facial features such as
\emph{cheek raise}, \emph{eye widen} or \emph{jaw drop}.

For both human and automatic recognition, to determine which of the six
``basic'' emotions can be used to identify each video, the recognition
probability for each label by frame was converted into odd ratio by
label {[}22{]}. The highest sum of each odd ratio time-series defines
the label recognized by the automatic classifier (EQ\ref{eq:2} for human
recognition and EQ\ref{eq:3} for automatic recognition).
\begin{equation}
\label{eq:2}
video_{i}.label = \max\left(\frac{\sum_{k=1}^{n}x_{video_{i}.label_{j}.t_{k}}}{\sum_{k=1}^{n}x_{video_{i}.t_{k}}}\right)
\end{equation}

@@

\hypertarget{iii.-results}{%
\section{III. Results}\label{iii.-results}}

@@

\hypertarget{a.-human-observers-accuracy}{%
\subsection{A. Human Observers'
Accuracy}\label{a.-human-observers-accuracy}}

The overall correlation of recognition and non-recognition between
self-reported emotions and human observers recognition is significant
but low (\(r = .24\), 95\% CI \([.19\), \(.29]\), \(t(1384) = 9.21\),
\(p < .001\)). In order to identify differences according the emotional
labels, encoders' subjective feeling is compared with human observers
recognition in a confusion matrix (Figure
\ref{fig:confusionMatrix_sr_hr}).

@@

Each emotion label used to describe encoders' self-reported subjective
feeling (\emph{i.e.}, the label rated with the highest value) is
compared with the emotion labels which were rated with the highest score
by human observers. Results of the confusion matrix show a moderate
agreement between the emotion felt by the encoder during the elicitation
and the emotion recognized by the human observers (Accuracy
\nolinebreak = \nolinebreak 0.43, 95\% CI {[}0.35,0.52{]}; Kappa = 0.19)
except for \emph{disgust} (100\% of the videos self-reported). These
results are far from those classicaly obtained in the literature for
emotional facial expression recognition which ranges between 60\% and
80\% accuracy. However these results are mostly obtained with static
(\emph{i.e.,} pictures) and posed (\emph{i.e.,} displayed by actors)
facial expressions using only 6 emotional labels in a forced-choice
paradigm.

Interestingly human observers seem to recognize \emph{surprise}
expressed in videos where \emph{anger}, \emph{fear} \emph{happiness} and
\emph{sadness} was the highest self-reported emotion (respectively
28.6\%, 33.3\%, 34.7\% and 33.3\% of the videos self-reported), and in a
lower instance \emph{happiness} was recognized in videos where
\emph{fear} and \emph{surprise} was the highest self-reported emotion
(respectively 50.0\% and 25.6\% of the videos self-reported).

Sensitivity, specificity, precision and F1 score for each emotion
reveals that \emph{happiness} has the highest coherence ratio whereas
\emph{sadness} has the lowest coherence ratio between true positives and
false positives (Table \ref{table:confusionTable_sr_hr}).

@@

Accuracy metrics by emotional labels indicate a discrepancy in the ratio
of true/false positives. Whereas \emph{happiness} and \emph{disgust}
obtain the highest scores, \emph{anger}, \emph{surprise} and
\emph{sadness} have the lowest recognition ratio. The underlying effect
of expression intensity may explain why \emph{happiness} and
\emph{disgust} are easily recognized. \emph{Anger} and \emph{sadness} as
non-socialy desirable emotion may be have been felt but not expressed.

@@

\hypertarget{b.-automatic-classifiers-accuracy}{%
\subsection{B. Automatic Classifier's
Accuracy}\label{b.-automatic-classifiers-accuracy}}

Similarly to the previous analysis, the overall correlation of
recognition and non-recognition revealed a significant but very low link
between self-reported emotions and automatic \nolinebreak classifier's
recognition (\(r = .12\), 95\% CI \([.07\), \(.17]\),
\(t(1384) = 4.50\), \(p < .001\)). A confusion matrix was used to
compare encoders' subjective feeling with the emotion label recognized
by the automatic classifier (Figure \ref{fig:confusionMatrix_sr_ar}).

@@

Results obtained for the comparison between emotions self-reported and
recognized by the automatic classifier are somewhat similar to the ones
with human observers (Table
\nolinebreak \ref{table:confusionTable_sr_ar}). Overall, a low agreement
between emotion self-reported and emotion recognized by the automatic
classifier (Accuracy \nolinebreak = \nolinebreak 0.3, 95\% CI
{[}0.22,0.38{]}; Kappa \nolinebreak = \nolinebreak 0.07) except for
\emph{happiness} (44.4\% of the video self-reported) is watched.

Surprisingly the automatic classifier incorrectly recognized as
\emph{disgust} an important proportion of videos in which \emph{anger},
\emph{happiness} and \emph{surprise} was the highest self-reported
emotion (respectively 42.9\%, 25.0\% and 32.6\% of the videos
self-reported). In parallel, the automatic classifier recognized as
\emph{happiness} videos in which \emph{fear} and \emph{surprise} was the
highest self-reported emotion (respectively 66.7\% and 32.6\% of the
videos self-reported).

@@

\hypertarget{c.-comparison-between-human-and-automatic-recognition}{%
\subsection{C. Comparison Between Human and Automatic
Recognition}\label{c.-comparison-between-human-and-automatic-recognition}}

As previously mentioned, human observers appear to be more accurate than
the automatic classifier to recognize individuals' subjective feeling
(human observers Accuracy \nolinebreak = \nolinebreak0.43; automatic
classifier Accuracy = 0.3; \(r = .22\), 95\% CI \([.17\), \(.27]\),
\(t(1384) = 8.31\), \(p < .001\)). However, both make mistakes.

A third confusion matrix is used to compare similarities (diagonal) and
differences between human observers and automatic classifier in
classifing the six emotion labels (Figure
\nolinebreak \ref{fig:confusionMatrix_hr_ar}).

@@

\hypertarget{iv.-discussion-and-conclusion}{%
\section{IV. Discussion and
Conclusion}\label{iv.-discussion-and-conclusion}}

@@

Such results support the hypothesis advanced by some authors of a low
emotion--expression coherence {[}23{]}. In many instances, facial
displays are not associated with a concordant emotional state, even any
emotional state at all {[}24{]}, {[}25{]}. More and more evidences are
showing that facial expressions are in reality not expressing emotions
{[}26{]}. As well as other nonverbal behaviors, facial movements are not
only assumed to be determined by emotion but also by various other
causes, such as psychological states (\emph{e.g.},
\nolinebreak motivations or pain), to say nothing of social context and
sociocultural norms {[}7{]}. This multiple determination excludes any
possibility of drawing an inference from facial activity on the
underlying psychological state (emotional or other).

Beyond the present observations showing a weak coherence between
subjective feelings and spontaneous facial expressions, this study
provides some light on the controversy between the behaviorist and the
constructivist approaches. The first of those two lines of thinking
{[}3{]} assumes that expressions of emotion are brief and coherent
patterns of facial muscle movements that co-vary with discrete
subjective experiences. Emotions and their related prototypical facial
behavior are universal because they are considered as innate mechanisms
allowing individuals to respond adaptively to evolutionary significant
events (threats, opportunities\ldots{}) encountered in the environment.
Instead of viewing emotions as natural kinds {[}12{]}, the
constructivist approach supposes that emotions are social constructions.
The specific emotion categories identified on the face of another person
are the result of categorization. In other words, the emotions that are
recognized by the observer are constructed in her/his mind, primarily
based on her/his conceptual knowledge of emotion. Therefore, facial
movements do not express specific emotions. It is the observer that
infers the emotional meaning of the facial expression. As a consequence,
one can predict from the first line of thinking that individuals'
emotional subjective feeling should be correlated to the recognition of
facial expressions from both human observers and automatic classifiers
whereas if emotions are social constructs, as stated by the second line
of thinking, human observer's should be better at perceiving emotions in
face than automatic classifiers.

Present results plead in favor of the latter stance. They show that
human observers are more accurate than the automatic recognition tool to
identify individuals' subjective feeling on the basis of their face.
Moreover, mistakes made by human observers look less arbitrary to the
ones made by the classifier. For instance, even if mix-up between
disgust and anger is sometimes reported in recognition studies,
confusions such as the ones produced by the classifier have never been
noted for human observers. A possible explanation is that human
observers are assessing dynamic expressive sequences. They have access
to the full video which provides a kinetic context whereas the automatic
classifier only assesses the videos frame by frame. Further research is
needed to shed light on this issue.

Some limitations should be stated, notably regarding the use of
self-reports to evaluate encoders' subjective feelings. Accessing the
inner subjective feeling can be biased if not impossible. Moreover the
procedure used for human observation can also be open to dispute.
Instead of asking the human annotators to provide a unique label, a more
subtle approach was chosen to mimic results provided by the automatic
classifier. Whereas this paradigm is longer and more complicated, it can
lead to more robust results in reducing the forced-choice bais {[}27{]}.
However this procedure can also reduce the human observers accuracy. In
this regard, the results of the human observation could have been more
ambiguous because it is not the natural way that people are inferring
meaning from facial expressions. An alternative explanation relies in
reducing the recognition bais involved in the classic recognition
paradigm. Classic forced-choice paradigms obtain artificial high
results, thus by using a more evolved approach observers' accuracy may
have been lowered.

Considering the above, the results provide additional evidence that
individuals' subjective feeling can not be inferred from facial
expressions and in our case invalidate the hypothesis of hardwired
emotions unambiguously displayed on the face. Even if emotions were
hardwired, in everyday life one does not observe prototypical facial
expressions and therefore research should be focused on analysing
non-prototypical facial expressions. Advancements in identifying
``non-basic'' emotion labels {[}28{]} as well as non-prototypical facial
expression have been made in the developpement of automatic facial
expression recognition tools. However this result suggests that
automatic facial expression recognition tools should merely evaluate
facial morphology features such as action units {[}29{]} rather than
inferring supposedly emotional or affective states.

\hypertarget{acknowledgment}{%
\section{V. Acknowledgment}\label{acknowledgment}}

@@

\hypertarget{references}{%
\section*{VI. References}\label{references}}
\addcontentsline{toc}{section}{VI. References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-dupre2018accuracy}{}%
{[}1{]} D. Dupré, N. Andelic, G. Morrison, and G. McKeown, ``Accuracy of
three commercial automatic emotion recognition systems across different
individuals and their facial expressions,'' in \emph{Proceedings of the
international conference on pervasive computing and communications},
2018, pp. 627--632.

\leavevmode\hypertarget{ref-darwin1872expression}{}%
{[}2{]} C. Darwin, \emph{The expression of the emotions in man and
animals}. London, UK: John Murray, 1872.
